---
title: "`r params$title`"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
  pdf_document:
    toc: true
  word_document: default
params:
  result_path: ""
  title: "GAM Model Report"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 5)
library(mgcvUI)
library(ggplot2)
gam_result <- readRDS(params$result_path)
model <- gam_result$model
summ <- format_gam_summary(gam_result)
```

## Model Overview

| Metric | Value |
|--------|-------|
| R-squared | `r round(summ$r_squared, 4)` |
| CV R-squared | `r if (!is.null(summ$cv_rsq)) round(summ$cv_rsq, 4) else "N/A"` |
| Deviance Explained | `r paste0(round(summ$dev_explained * 100, 1), "%")` |
| AIC | `r round(summ$aic, 1)` |
| BIC | `r round(summ$bic, 1)` |
| Observations | `r summ$n_obs` |
| Smooth Terms | `r summ$n_smooths` |
| Method | `r summ$method` |
| Family | `r summ$family` |
| Fitting Time | `r round(gam_result$elapsed, 2)` sec |

## Formula

```
`r paste(deparse(gam_result$formula, width.cutoff = 80), collapse = "\n")`
```

## Summary

### Smooth Terms

```{r smooth-terms}
if (nrow(summ$smooth_table) > 0) {
  st <- summ$smooth_table
  st$EDF <- round(st$EDF, 2)
  st$Ref.df <- round(st$Ref.df, 2)
  st$F <- round(st$F, 2)
  st$p_value <- format.pval(st$p_value, digits = 3)
  knitr::kable(st)
}
```

### Parametric Terms

```{r parametric-terms}
if (nrow(summ$parametric_table) > 0) {
  pt <- summ$parametric_table
  pt$Estimate <- round(pt$Estimate, 4)
  pt$Std_Error <- round(pt$Std_Error, 4)
  pt$t_value <- round(pt$t_value, 3)
  pt$p_value <- format.pval(pt$p_value, digits = 3)
  knitr::kable(pt)
}
```

## Variable Importance

```{r importance-plot, fig.height=4}
imp_df <- data.frame(Term = character(0), Statistic = numeric(0),
                     Type = character(0), stringsAsFactors = FALSE)

if (nrow(summ$smooth_table) > 0) {
  imp_df <- rbind(imp_df, data.frame(
    Term = summ$smooth_table$Term,
    Statistic = summ$smooth_table$F,
    Type = "Smooth", stringsAsFactors = FALSE))
}

if (nrow(summ$parametric_table) > 0) {
  pt <- summ$parametric_table[summ$parametric_table$Term != "(Intercept)", ,
                               drop = FALSE]
  if (nrow(pt) > 0) {
    imp_df <- rbind(imp_df, data.frame(
      Term = pt$Term, Statistic = abs(pt$t_value),
      Type = "Parametric", stringsAsFactors = FALSE))
  }
}

if (nrow(imp_df) > 0) {
  imp_df$Term <- factor(imp_df$Term,
                        levels = imp_df$Term[order(imp_df$Statistic)])
  ggplot(imp_df, aes(x = .data$Statistic, y = .data$Term,
                     fill = .data$Type)) +
    geom_col(alpha = 0.85) +
    scale_fill_manual(values = c("Smooth" = "#5e81ac",
                                 "Parametric" = "#a3be8c")) +
    labs(title = "Variable Importance (F / |t| statistic)",
         x = "Statistic", y = NULL) +
    theme_minimal(base_size = 13)
}
```

```{r importance-table}
if (nrow(imp_df) > 0) {
  tbl <- imp_df[order(imp_df$Statistic, decreasing = TRUE), ]
  tbl$Statistic <- round(tbl$Statistic, 4)
  knitr::kable(tbl, row.names = FALSE)
}
```

## Smooth Term Plots

```{r smooths, fig.width=7, fig.height=4, results='asis'}
for (spec in gam_result$smooth_specs) {
  if (spec$type != "s" || length(spec$vars) != 1L) next
  p <- tryCatch(
    plot_smooth_single(gam_result, spec$vars[1],
                       earth_knots = gam_result$earth_knots),
    error = function(e) NULL
  )
  if (!is.null(p)) {
    cat("\n###", spec$vars[1], "\n\n")
    print(p)
    cat("\n")
  }
}
```

## Correlation Matrix

```{r correlation, fig.width=7, fig.height=7}
mf <- model$model
pred_data <- mf[, setdiff(names(mf), gam_result$response), drop = FALSE]
num_cols <- vapply(pred_data, is.numeric, logical(1))
pred_data <- pred_data[, num_cols, drop = FALSE]

if (ncol(pred_data) >= 2L) {
  cor_mat <- cor(pred_data, use = "pairwise.complete.obs")
  vars <- colnames(cor_mat)
  cor_long <- expand.grid(Var1 = vars, Var2 = vars, stringsAsFactors = FALSE)
  cor_long$value <- as.numeric(cor_mat)
  cor_long$Var1 <- factor(cor_long$Var1, levels = rev(vars))
  cor_long$Var2 <- factor(cor_long$Var2, levels = vars)
  n_vars <- length(vars)
  txt_size <- max(2.5, 5 - n_vars * 0.15)

  ggplot(cor_long, aes(x = .data$Var2, y = .data$Var1,
                       fill = .data$value)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = sprintf("%.2f", .data$value)), size = txt_size) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                         midpoint = 0, limits = c(-1, 1),
                         name = "Correlation") +
    coord_fixed() +
    labs(title = "Correlation Matrix", x = NULL, y = NULL) +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank())
}
```

## Diagnostics

```{r diagnostics, fig.width=8, fig.height=8}
plot_diagnostics(gam_result)
```

## Actual vs Predicted

```{r avp, fig.width=6, fig.height=5}
plot_actual_vs_predicted(gam_result)
```

## ANOVA

```{r anova}
aov <- summary(model)
parts <- list()

if (!is.null(aov$p.table) && nrow(aov$p.table) > 0) {
  pt <- as.data.frame(aov$p.table)
  pt$Term <- rownames(pt)
  pt$Type <- "parametric"
  parts <- c(parts, list(pt))
}

if (!is.null(aov$s.table) && nrow(aov$s.table) > 0) {
  st <- as.data.frame(aov$s.table)
  st$Term <- rownames(st)
  st$Type <- "smooth"
  parts <- c(parts, list(st))
}

if (length(parts) > 0) {
  all_cols <- unique(unlist(lapply(parts, names)))
  aov_df <- do.call(rbind, lapply(parts, function(p) {
    missing <- setdiff(all_cols, names(p))
    for (m in missing) p[[m]] <- NA
    p[, all_cols, drop = FALSE]
  }))
  front <- c("Term", "Type")
  aov_df <- aov_df[, c(front, setdiff(names(aov_df), front)), drop = FALSE]
  rownames(aov_df) <- NULL
  pcol <- grep("p-value|Pr\\(", names(aov_df), value = TRUE)
  for (pc in pcol) aov_df[[pc]] <- format.pval(aov_df[[pc]], digits = 3)
  knitr::kable(aov_df, digits = 4)
}
```

## Concurvity

### Overall

```{r concurvity-full}
if (length(model$smooth) > 0L) {
  con <- tryCatch(mgcv::concurvity(model, full = TRUE), error = function(e) NULL)
  if (!is.null(con)) {
    con_df <- as.data.frame(round(con, 4))
    knitr::kable(con_df)
  }
}
```

### Pairwise (worst case)

```{r concurvity-pairwise}
if (length(model$smooth) >= 2L) {
  con_pw <- tryCatch(mgcv::concurvity(model, full = FALSE),
                     error = function(e) NULL)
  if (!is.null(con_pw) && is.list(con_pw) && "worst" %in% names(con_pw)) {
    pw_df <- as.data.frame(round(con_pw$worst, 4))
    knitr::kable(pw_df)
  }
}
```

```{r sign-check, results='asis'}
signs <- check_sign_consistency(gam_result)
if (!is.null(signs)) {
  cat("\n## Earth vs GAM Sign Consistency\n\n")
  knitr::kable(signs[, c("variable", "earth_direction",
                         "gam_direction", "consistent")])
}
```

## Basis Dimension Check

```{r gam-check}
mgcv::gam.check(model)
```

## Mgcv Output

```{r mgcv-output, comment=""}
cat(sprintf("== Timing: %.2f seconds ==\n\n", gam_result$elapsed))
cat("== Formula ==\n\n")
print(gam_result$formula)
cat("\n\n== Summary ==\n\n")
print(summary(model))
cat("\n\n== Family ==\n\n")
print(model$family)
```

---

*Report generated by mgcvUI on `r Sys.time()`*
