---
title: "GAM Model Report"
format:
  html:
    toc: true
    theme: flatly
    html-math-method: katex
  pdf:
    toc: true
    papersize: letter
    include-in-header:
      text: |
        \usepackage{longtable}
        \usepackage{booktabs}
  docx:
    toc: true
    reference-doc: reference.docx
params:
  data_file: ""
  paper_size: "letter"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
`%||%` <- function(a, b) if (is.null(a)) b else a

# Resolve the assets payload: explicit param wins; else a sibling
# report_data.rds (self-contained bundle), so `quarto render` works with no
# external arguments.
.data_file <- if (nzchar(params$data_file) && file.exists(params$data_file)) {
  params$data_file
} else if (file.exists("report_data.rds")) {
  "report_data.rds"
} else {
  ""
}

.assets <- if (nzchar(.data_file)) readRDS(.data_file) else NULL
result       <- .assets$result
summary_info <- .assets$summary_info
importance   <- .assets$importance
equation       <- .assets$equation
equation_latex <- .assets$equation_latex
equation_defs  <- .assets$equation_defs
summary_text   <- .assets$summary_text

.plots_dir <- if (nzchar(.data_file)) file.path(dirname(.data_file), "plots") else "plots"
if (!dir.exists(.plots_dir) && dir.exists("plots")) .plots_dir <- "plots"
.has_assets <- dir.exists(.plots_dir) && length(list.files(.plots_dir)) > 0L

fmt <- if (!is.null(knitr::pandoc_to())) knitr::pandoc_to() else "html"
.plot_ext <- if (fmt == "latex") "pdf" else "png"

# Include a pre-generated plot if it exists. For PDF use the relative path;
# for HTML/Word embed as a data URI so the image travels with the file.
include_plot <- function(name) {
  f <- file.path(.plots_dir, paste0(name, ".", .plot_ext))
  if (!file.exists(f)) return(invisible(NULL))
  if (fmt == "latex") {
    cat(sprintf("![](%s)\n\n", file.path("plots", paste0(name, ".", .plot_ext))))
  } else {
    cat(sprintf("![](%s)\n\n", knitr::image_uri(f)))
  }
  invisible(NULL)
}
```

## Dataset & Model

| Property | Value |
|----------|-------|
| Response | `r result$response` |
| Predictors | `r paste(result$predictors, collapse = ", ")` |
| Observations | `r summary_info$n_obs` |
| Family | `r summary_info$family` |
| Smoothing method | `r summary_info$method` |
| Smooth terms | `r summary_info$n_smooths` |

## Results Summary

| Metric | Value |
|--------|-------|
| R-squared (adj) | `r sprintf("%.4f", summary_info$r_squared)` |
| Deviance explained | `r sprintf("%.1f%%", 100 * summary_info$dev_explained)` |
| `r if (!is.null(summary_info$cv_rsq)) "CV R-squared" else "CV R-squared (n/a)"` | `r if (!is.null(summary_info$cv_rsq)) sprintf("%.4f", summary_info$cv_rsq) else "—"` |
| AIC | `r sprintf("%.1f", summary_info$aic)` |
| BIC | `r if (is.finite(summary_info$bic)) sprintf("%.1f", summary_info$bic) else "—"` |

## Model Equation

```{r equation, results='asis'}
if (!is.null(equation_latex) && nzchar(equation_latex)) {
  # Display math, one term per line (aligned) so it never runs off the page.
  cat("\n$$\n", equation_latex, "\n$$\n\n", sep = "")
  if (length(equation_defs)) {
    cat("**Smooth function definitions:**\n\n")
    for (d in equation_defs) {
      cat(sprintf("- $%s$ → `%s` — %s\n", d$fn, d$label, d$desc))
    }
    cat("\n")
  }
} else {
  # Fallback: raw R formula in a code block.
  cat("```\n", equation, "\n```\n", sep = "")
}
```

## Smooth Terms

```{r smooth-table}
st <- summary_info$smooth_table
if (nrow(st) > 0L) {
  st$EDF <- round(st$EDF, 3); st$Ref.df <- round(st$Ref.df, 3)
  st$F <- round(st$F, 3)
  st$p_value <- format.pval(st$p_value, digits = 3, eps = 1e-4)
  knitr::kable(st)
} else cat("_No smooth terms._\n")
```

## Parametric Terms

```{r param-table}
pt <- summary_info$parametric_table
if (nrow(pt) > 0L) {
  pt$Estimate <- signif(pt$Estimate, 4); pt$Std_Error <- signif(pt$Std_Error, 4)
  pt$t_value <- round(pt$t_value, 3)
  pt$p_value <- format.pval(pt$p_value, digits = 3, eps = 1e-4)
  knitr::kable(pt)
} else cat("_No parametric terms._\n")
```

## Variable Importance

```{r importance, results='asis'}
include_plot("importance")
```

## Smooth Effects

```{r smooths, results='asis'}
# Smooth panels are split into page-sized chunks so each renders large.
nch <- .assets$n_smooth_chunks
if (is.null(nch) || nch < 1) {
  include_plot("smooths")            # backward-compat (older bundles)
} else {
  for (i in seq_len(nch)) include_plot(paste0("smooths_", i))
}
```

```{r per-term, results='asis'}
specs <- result$smooth_specs
if (!is.null(specs)) {
  for (i in seq_along(specs)) {
    sp <- specs[[i]]
    if (identical(sp$type %||% "s", "linear")) next
    lbl <- paste(sp$vars, collapse = " × ")
    cat(sprintf("\n### %s\n\n", lbl))
    include_plot(paste0("term_", i))
  }
}
```

## Predictor Correlations

```{r correlation, results='asis'}
include_plot("correlation")
```

## Diagnostics

```{r diagnostics, results='asis'}
include_plot("diagnostics")
```

### Actual vs Predicted

```{r avp, results='asis'}
include_plot("actual_vs_predicted")
```

## mgcv Summary

```{r raw-summary, comment=""}
cat(summary_text)
```

---

*Report generated by mgcvUI on `r format(Sys.time())`.*
