---
title: "Using earthUI and mgcvUI Together"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using earthUI and mgcvUI Together}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The Two-Stage Workflow

Real estate valuation models benefit from a two-stage approach:

1. **Discover structure** with `earthUI` — MARS (Multivariate Adaptive
   Regression Splines) is excellent at finding breakpoints (knots) and
   identifying which variables matter.
2. **Confirm and report** with `mgcvUI` — GAMs provide smooth,
   statistically rigorous curves with proper confidence intervals and
   p-values, which are better suited for defensible appraisal reports.

The idea: let earth find the knots, then let mgcv smooth through them.

## Step 1: Fit an Earth Model in earthUI

```{r eval=FALSE}
library(earthUI)
earthUI::launch()
```

In the earthUI app:

1. Upload your sales data.
2. Select sale price as the response, property features as predictors.
3. Fit the model.
4. **Save the result** — click the download button to save the
   earthUI result as an `.rds` file.

Or programmatically:

```{r eval=FALSE}
library(earthUI)
result <- earthUI::fit_earth(
  df         = sales_data,
  target     = "sale_price",
  predictors = c("living_area", "lot_size", "year_built", "condition")
)
saveRDS(result, "earth_result.rds")
```

## Step 2: Import into mgcvUI

```{r eval=FALSE}
library(mgcvUI)
mgcvUI()
```

In the mgcvUI app:

1. Upload your sales data on the **Data** tab.
2. Go to the **Earth Import** tab and upload the `.rds` file.
3. mgcvUI extracts the knot positions and pre-configures the GAM:
   - Predictors are auto-selected to match the earth model.
   - Basis dimensions (`k`) are set based on the number of earth knots.
   - Cubic regression splines (`cr`) are used by default for variables
     with earth knots, which allows the knot positions to be passed to
     `gam()`.
4. Fit the model on the **Model** tab.
5. Check the **Sign Check** sub-tab to verify that the GAM smooth
   directions agree with the earth hinge directions.

## Programmatic Pipeline

```{r eval=FALSE}
library(mgcvUI)

# Import earth result and extract knots
ek <- import_earth("earth_result.rds")

# View the knots
str(ek$knots)

# Build GAM specs informed by earth knots
specs <- list(
  list(vars = "living_area", type = "s", bs = "cr",
       k = length(ek$knots$living_area) + 2L),
  list(vars = "lot_size", type = "s", bs = "cr",
       k = length(ek$knots$lot_size) + 2L),
  list(vars = "year_built", type = "s", bs = "cr",
       k = length(ek$knots$year_built) + 2L),
  list(vars = "condition", type = "linear", bs = NULL, k = NULL)
)

# Fit GAM with earth knots
result <- fit_gam(
  data        = sales_data,
  response    = "sale_price",
  smooth_specs = specs,
  earth_knots = ek
)

# Check sign consistency
check_sign_consistency(result)

# Export knots to CSV for documentation
export_knots_csv(ek, "knot_positions.csv")

# Generate appraisal report
render_gam_report(result, "docx", "appraisal_model_report.docx",
                  title = "GAM Valuation Model — Main Street Subdivision")
```

## Understanding Sign Consistency

When you import earth knots, mgcvUI checks whether the GAM smooth
curves agree with the earth hinge directions:

| Earth Direction | Meaning |
|----------------|---------|
| Forward (+) | The hinge function increases after the knot — earth implies a positive (increasing) relationship in that region |
| Reverse (-) | The hinge function decreases after the knot — earth implies a negative (decreasing) relationship |

The **Sign Check** tab compares these with the overall direction of the
fitted GAM smooth. A **consistent** result means the GAM and earth
agree. An **inconsistent** result flags variables where the GAM smooth
goes in the opposite direction from what earth found — this may indicate:

- The relationship is genuinely complex (U-shaped, etc.)
- The earth model over-fit in that region
- Additional predictors are needed to account for confounding

Inconsistency is a flag for investigation, not necessarily a problem.

## Exporting Knots

The knot CSV contains one row per knot with columns:

| Column | Description |
|--------|-------------|
| `variable` | Predictor name |
| `knot` | The knot value (breakpoint) |
| `direction` | `"forward"` or `"reverse"` |

This provides a transparent record of the MARS-derived structure that
informed the GAM, suitable for inclusion in appraisal workfiles.
