---
title: "Getting Started with mgcvUI for Real Estate Appraisers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mgcvUI for Real Estate Appraisers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is mgcvUI?

mgcvUI is an interactive tool for building **Generalized Additive Models
(GAMs)** — flexible regression models that can capture nonlinear
relationships in your data without requiring you to specify the exact
shape of each curve.

If you are a real estate appraiser, you already know that the
relationship between sale price and living area is not a straight line:
the per-square-foot value often changes as homes get larger. A GAM can
learn these curves directly from your data, producing a transparent,
defensible model.

## Quick Start

### 1. Launch the app

```{r eval=FALSE}
library(mgcvUI)
mgcvUI()
```

This opens mgcvUI in your web browser.

### 2. Import your data

On the **Data** tab, upload a CSV or Excel file containing your
comparable sales. mgcvUI expects one row per sale with columns for the
sale price and property characteristics (living area, lot size, age,
etc.).

### 3. Choose your variables

On the **Variables** tab:

- **Response** — select your sale price column.
- **Predictors** — select the property characteristics you want in the
  model.

mgcvUI automatically configures a smooth curve for each numeric
predictor. Categorical variables (e.g. neighborhood, condition) enter
the model as linear terms.

### 4. Fit the model

On the **Model** tab, click **Fit Model**. mgcvUI fits a GAM using
`mgcv::gam()` with the REML method (the statistical gold standard for
choosing how wiggly each curve should be).

### 5. Review the results

The **Model** tab has sub-tabs for:

- **Summary** — R-squared, AIC, and significance of each term.
- **Smooth Plots** — visual curves for each predictor, with confidence
  bands showing statistical uncertainty.
- **Diagnostics** — residual plots and Q-Q plots to check model
  assumptions.
- **Basis Check** — verifies that the model has enough flexibility to
  capture the patterns in your data.

### 6. Generate a report

On the **Report** tab, download a formatted report (HTML, Word, or PDF)
to include in your appraisal workfile.

## Understanding the Smooth Plots

Each smooth plot shows the estimated partial effect of one predictor on
the sale price, holding all other predictors constant:
- The **blue line** is the estimated curve.
- The **blue band** is the 95% confidence interval.
- The **grey dots** are partial residuals (data points adjusted for the
  effects of other predictors).

A flat curve means the predictor has little effect. A steep curve means
the predictor is influential.

## Adjusting Mgcv Call Parameters

Most users can leave the defaults. Power users may want to adjust:

- **Basis type** — `tp` (thin plate, the default) is an excellent
  all-rounder. `cr` (cubic regression spline) is slightly faster and
  works well when you have earth knot positions to use.
- **k** — the basis dimension controls the maximum wiggliness. Higher
  values allow more complex curves. The default (`auto`) lets mgcv
  choose.
- **Gamma** — values greater than 1 produce smoother curves (useful if
  you suspect over-fitting).
- **Variable selection** — when enabled, terms that are not needed are
  penalised toward zero.

## Programmatic Use

You do not need the Shiny app. The core functions work in plain R:

```{r}
library(mgcvUI)

specs <- list(
  list(vars = "wt", type = "s", bs = "tp", k = NULL),
  list(vars = "hp", type = "s", bs = "tp", k = NULL)
)

result <- fit_gam(mtcars, "mpg", specs)
format_gam_summary(result)
```

```{r fig.width=7, fig.height=4}
plot_smooths(result)
```

```{r fig.width=6, fig.height=5}
plot_actual_vs_predicted(result)
```
