---
title: "`r params$title`"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: flatly
  pdf_document:
    toc: true
  word_document: default
params:
  result_path: ""
  title: "GAM Model Report"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(mgcvUI)
library(ggplot2)
gam_result <- readRDS(params$result_path)
model <- gam_result$model
summ <- format_gam_summary(gam_result)
```

## Model Overview

| Metric | Value |
|--------|-------|
| R-squared | `r round(summ$r_squared, 4)` |
| Deviance Explained | `r paste0(round(summ$dev_explained * 100, 1), "%")` |
| AIC | `r round(summ$aic, 1)` |
| BIC | `r round(summ$bic, 1)` |
| Observations | `r summ$n_obs` |
| Smooth Terms | `r summ$n_smooths` |
| Method | `r summ$method` |
| Family | `r summ$family` |
| Fitting Time | `r round(gam_result$elapsed, 2)` sec |

## Formula

```
`r deparse(gam_result$formula, width.cutoff = 80)`
```

## Summary

```{r summary}
summary(model)
```

## Smooth Term Plots

```{r smooths, fig.width=8, fig.height=6}
plot_smooths(gam_result)
```

## Diagnostics

```{r diagnostics, fig.width=8, fig.height=8}
plot_diagnostics(gam_result)
```

## Actual vs Predicted

```{r avp, fig.width=6, fig.height=5}
plot_actual_vs_predicted(gam_result)
```

```{r sign-check, results='asis'}
signs <- check_sign_consistency(gam_result)
if (!is.null(signs)) {
  cat("\n## Earth vs GAM Sign Consistency\n\n")
  knitr::kable(signs[, c("variable", "earth_direction",
                         "gam_direction", "consistent")])
}
```

## Basis Dimension Check

```{r gam-check}
mgcv::gam.check(model)
```
