---
title: "Using earthUI and mgcvUI Together"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using earthUI and mgcvUI Together}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The Two-Stage Workflow

Real estate valuation models benefit from a two-stage approach:

1. **Discover structure** with `earthUI` -- MARS (Multivariate Adaptive
   Regression Splines) is excellent at finding breakpoints (knots) and
   identifying which variables matter.
2. **Confirm and report** with `mgcvUI` -- GAMs provide smooth,
   statistically rigorous curves with proper confidence intervals and
   p-values, which are better suited for defensible appraisal reports.

The idea: let earth find the knots, then let mgcv smooth through them.

## What mgcvUI Extracts from earthUI

When you import an earthUI result `.rds` file, mgcvUI extracts the
following components:

| Component | Source in Earth Model | How mgcvUI Uses It |
|-----------|----------------------|---------------------|
| **Knot positions** | `model$cuts` + `model$selected.terms` | Placed as `cr` basis knots via `gam(..., knots=)`. Earth found the breakpoints; mgcv smooths through them |
| **Hinge directions** | `model$dirs` (+1 = forward, -1 = reverse) | Sign consistency check compares earth direction with GAM smooth derivative to flag contradictions |
| **Interactions** | Rows in `dirs` where 2+ predictors nonzero | Pre-checks the Allowed Interactions matrix |
| **Target variable** | `earth_result$target` | Auto-sets the response dropdown |
| **Predictors** | `earth_result$predictors` | Auto-checks "Inc" for each predictor |
| **Categoricals** | `earth_result$categoricals` | Auto-checks "Factor" |
| **Linear predictors** | `earth_result$linpreds` | Auto-checks "Linear" |
| **Interaction degree** | `earth_result$degree` | Info message if degree=1 |
| **Allowed matrix** | `earth_result$allowed_matrix` | Seeds interaction checkboxes |
| **R-sq, GCV, n_terms** | `model$rsq`, `model$gcv` | Displayed in import summary |

The key insight: **earth's job is discovery, mgcv's job is regularized
estimation**. The knots are the bridge -- they tell mgcv *where* the
nonlinearities are so it doesn't have to find them itself.

## Parameter Presets: Why They Differ

When an earthUI result is imported, mgcvUI auto-switches to the
**Earth Pipeline** preset:

| Setting | Standalone | Earth Pipeline | Why Different |
|---------|-----------|----------------|---------------|
| Basis | `tp` | `cr` | Earth provides knot positions; `cr` uses them directly |
| k | auto (10) | 5 | Earth already found breakpoints; low k prevents refitting |
| Gamma | 1.2 | 1.4 | Stronger parsimony -- mgcv refines, not discovers |
| Select | TRUE | TRUE | Second-stage shrinkage guard |

Every setting reinforces the division: earth discovers, mgcv estimates.

## Step 1: Fit an Earth Model in earthUI

```{r eval=FALSE}
library(earthUI)
earthUI::launch()
```

In the earthUI app:

1. Upload your sales data.
2. Select sale price as the response, property features as predictors.
3. Fit the model.
4. **Save the result** -- click the download button to save the
   earthUI result as an `.rds` file.

Or programmatically:

```{r eval=FALSE}
library(earthUI)
result <- earthUI::fit_earth(
  df         = sales_data,
  target     = "sale_price",
  predictors = c("living_area", "lot_size", "year_built", "condition")
)
saveRDS(result, "earth_result.rds")
```

## Step 2: Import into mgcvUI

```{r eval=FALSE}
library(mgcvUI)
mgcvUI()
```

In the mgcvUI app:

1. Upload your sales data via **Browse** in the Data section.
2. Expand **Import from earthUI** and browse for the `.rds` file.
   Files are read directly from disk -- no upload size limit.
3. mgcvUI extracts the knot positions and pre-configures the GAM:
   - Predictors are auto-selected to match the earth model.
   - Basis dimensions (`k`) are set based on the number of earth knots.
   - Cubic regression splines (`cr`) are used for variables with earth
     knots, allowing knot positions to be passed to `gam()`.
   - The parameter preset auto-switches to **Earth Pipeline**.
4. Review the **Predictor Settings** table -- earth's categoricals
   are flagged as Factor, linear predictors are flagged as Linear.
5. Fit the model.
6. Check the **Sign Check** tab to verify GAM smooth directions agree
   with earth hinge directions.
7. Check the **Concurvity** tab -- earth's hinge functions on correlated
   predictors can produce near-concurve smooth terms. Values > 0.8
   warrant collapsing or dropping terms.

## Concurvity: Why It Matters in This Pipeline

Earth's hinge functions on correlated predictors (GLA and room count,
for instance) can produce near-concurve smooth terms in mgcv -- the
smooth analog of multicollinearity.

The **Concurvity** tab shows:

- **Overall**: worst/observed/estimate for each smooth term
- **Pairwise**: worst-case between each pair of smooth terms

Color coding: white (< 0.5), yellow (0.5--0.8), red (> 0.8).

Values above ~0.8 in the "worst" row warrant collapsing or dropping one
of the terms before treating the model as defensible.

## Programmatic Pipeline

```{r eval=FALSE}
library(mgcvUI)

# Import earth result and extract knots
ek <- import_earth("earth_result.rds")

# View the knots
str(ek$knots)

# Build GAM specs informed by earth knots
specs <- list(
  list(vars = "living_area", type = "s", bs = "cr",
       k = length(ek$knots$living_area)),
  list(vars = "lot_size", type = "s", bs = "cr",
       k = length(ek$knots$lot_size)),
  list(vars = "year_built", type = "s", bs = "cr",
       k = length(ek$knots$year_built)),
  list(vars = "condition", type = "linear", bs = NULL, k = NULL)
)

# Fit GAM with earth knots -- note earth pipeline defaults
result <- fit_gam(
  data         = sales_data,
  response     = "sale_price",
  smooth_specs = specs,
  earth_knots  = ek,
  method       = "REML",
  select       = TRUE,
  gamma        = 1.4
)

# Check concurvity
mgcv::concurvity(result$model, full = TRUE)

# Check sign consistency
check_sign_consistency(result)

# Export knots to CSV for documentation
export_knots_csv(ek, "knot_positions.csv")

# Generate appraisal report
render_gam_report(result, "docx", "appraisal_model_report.docx",
                  title = "GAM Valuation Model")
```

## Understanding Sign Consistency

When you import earth knots, mgcvUI checks whether the GAM smooth
curves agree with the earth hinge directions:

| Earth Direction | Meaning |
|----------------|---------|
| Forward (+) | The hinge function increases after the knot -- earth implies a positive relationship in that region |
| Reverse (-) | The hinge function decreases after the knot -- earth implies a negative relationship |

The **Sign Check** tab compares these with the overall direction of the
fitted GAM smooth. A **consistent** result means the GAM and earth
agree. An **inconsistent** result flags variables where the GAM smooth
goes in the opposite direction from what earth found -- this may indicate:

- The relationship is genuinely complex (U-shaped, etc.)
- The earth model over-fit in that region
- Additional predictors are needed to account for confounding

Inconsistency is a flag for investigation, not necessarily a problem.

## Exporting Knots

The knot CSV contains one row per knot with columns:

| Column | Description |
|--------|-------------|
| `variable` | Predictor name |
| `knot` | The knot value (breakpoint) |
| `direction` | `"forward"` or `"reverse"` |

This provides a transparent record of the MARS-derived structure that
informed the GAM, suitable for inclusion in appraisal workfiles.
