---
title: "Getting Started with mgcvUI for Real Estate Appraisers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mgcvUI for Real Estate Appraisers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is mgcvUI?

mgcvUI is an interactive tool for building **Generalized Additive Models
(GAMs)** -- flexible regression models that can capture nonlinear
relationships in your data without requiring you to specify the exact
shape of each curve.

If you are a real estate appraiser, you already know that the
relationship between sale price and living area is not a straight line:
the per-square-foot value often changes as homes get larger. A GAM can
learn these curves directly from your data, producing a transparent,
defensible model.

## Prerequisites

mgcvUI works on macOS, Windows, and Linux with R >= 4.1.0. All features
work out of the box except:

- **PDF reports** require a LaTeX installation. Install with
  `tinytex::install_tinytex()`. If LaTeX is not detected, the PDF
  option is automatically hidden and HTML/Word reports remain available.
- **Roboto Condensed font** is downloaded from Google Fonts for plot
  styling. If your machine is offline, the app uses the system
  sans-serif font automatically.

## Quick Start

### 1. Launch the app

```{r eval=FALSE}
library(mgcvUI)
mgcvUI()
```

This opens mgcvUI in your web browser.

### 2. Create or open a project

In sidebar section 1, create or open a **project**. A project sets the
purpose (General, For Appraisal, or Market Area Analysis) and the work
location (country, state, county, city), and provides the input and
output folders. Projects live under a shared `~/regProj` root and are
shared with the sibling apps earthUI and glmnetUI. The purpose is fixed
when the project is created -- open or create a different project to
work in another mode.

### 3. Import your data

Drop your CSV or Excel file of comparable sales into the project's
`<os>_in/` folder (via Finder/Explorer), then pick it from the file list
in section 2 (click **Refresh** if it does not appear). Files are read
directly from disk (no upload size limits). Each project remembers its
last-used file and auto-loads it when the project is reopened.

### 4. Choose your variables

In the **Predictor Settings** table:

- **Response** -- select your sale price column.
- **Inc** -- check the predictors you want in the model.
- **Factor** -- check categorical variables (e.g., neighborhood, condition).
- **Linear** -- check variables that should enter linearly (no smooth curve).
- **Special** -- assign column roles like `living_area`, `lot_size`,
  `area`, `weight`, `sale_age`, `sale_type`, `latitude`, `longitude`, etc.

mgcvUI automatically configures a smooth curve for each included numeric
predictor. Categorical variables enter the model as linear terms.

### 5. Choose your parameter preset

The **Parameter Preset** selector at the top of Mgcv Call Parameters
offers two options:

- **Standalone (discovery)** -- for GAMs without earth input. Uses `tp`
  basis, k=auto(10), gamma=1.2. mgcv discovers the nonlinearities.
- **Earth Pipeline (refinement)** -- auto-selected when an earthUI
  result is imported. Uses `cr` basis, k=auto(10), gamma=1.4, with earth
  knots preserved and augmented with data quantiles. mgcv refines
  earth's breakpoints.

See the [mgcvUI User Guide](mgcvUI-user-guide.html) for full parameter
details.

### 6. Fit the model

Click **Fit Model**. mgcvUI fits a GAM using `mgcv::gam()` with your
chosen parameters and shows a notification with the fitting time.

### 7. Review the results

The results panel has tabs for:

- **Summary** -- R-squared, CV R-squared, AIC, and significance of each term.
- **Contribution** -- interactive smooth plots for each predictor, with
  confidence bands showing statistical uncertainty.
- **Concurvity** -- checks for near-concurvity (smooth multicollinearity).
  Values above 0.8 warrant investigation.
- **Diagnostics** -- residual plots and Q-Q plots to check model
  assumptions.
- **Basis Check** -- verifies that the model has enough flexibility.

### 8. Generate a report

Use **Download Output** to export estimated sale prices and residuals
as Excel. Reports use a two-step Quarto workflow: **Generate Quarto
Report** writes an editable `.qmd` report bundle to the project's output
folder, and **Convert Quarto Report** renders it to HTML, Word, or PDF
(PDF requires LaTeX) for your appraisal workfile. In appraisal mode you
can also compute RCA (Residual Constraint Approach) adjustments and a
Sales Comparison Grid.

## Understanding the Smooth Plots

Each smooth plot shows the estimated partial effect of one predictor on
the sale price, holding all other predictors constant:
- The **blue line** is the estimated curve.
- The **blue band** is the 95% confidence interval.
- The **grey dots** are partial residuals (data points adjusted for the
  effects of other predictors).

A flat curve means the predictor has little effect. A steep curve means
the predictor is influential.

## Understanding Concurvity

Concurvity is the smooth analog of multicollinearity. If two predictors
are highly correlated (e.g., living area and number of rooms), their
smooth terms can become nearly interchangeable, making individual effects
unreliable.

The **Concurvity** tab shows overall and pairwise concurvity values.
Cells are color-coded: white (< 0.5, OK), yellow (0.5--0.8, moderate),
red (> 0.8, high). High values mean you should consider dropping or
combining correlated terms.

## Adjusting Mgcv Call Parameters

Most users can use the preset defaults. Power users may want to adjust:

- **Basis type** -- `tp` (thin plate, the default for standalone) is an
  excellent all-rounder. `cr` (cubic regression spline) is faster and
  works well when you have earth knot positions.
- **k** -- the basis dimension controls maximum wiggliness. Higher
  values allow more complex curves. Too high risks overfitting on small
  comp sets.
- **Gamma** -- values greater than 1 produce smoother curves. 1.4 is the
  Wood & Fasiolo recommendation for parsimony.
- **Select** -- when enabled (recommended), terms that contribute little
  are shrunk toward zero. This is the `select` argument to `mgcv::gam()`.
- **Method** -- REML is strongly preferred over GCV.Cp for correlated
  predictors and modest sample sizes.

## Interactions

### Tensor Interactions (continuous x continuous)

Under "Allowed Interactions", the matrix shows pairs of smooth predictors.
Check a cell to add a `ti()` tensor interaction.

### Factor-by-Smooth Interactions

The "Factor-by-Smooth" grid lets you fit separate smooth curves per
factor level. For example, checking `lot_size` x `area_id` generates
`s(lot_size, by=area_id)` -- a different lot size curve for each neighborhood.

## Programmatic Use

You do not need the Shiny app. The core functions work in plain R:

```{r}
library(mgcvUI)

specs <- list(
  list(vars = "wt", type = "s", bs = "tp", k = NULL),
  list(vars = "hp", type = "s", bs = "tp", k = NULL)
)

result <- fit_gam(mtcars, "mpg", specs)
format_gam_summary(result)
```

```{r fig.width=7, fig.height=4}
plot_smooths(result)
```

```{r fig.width=6, fig.height=5}
plot_actual_vs_predicted(result)
```
